---
title: "quantify()-ing binary outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quantify()-ing binary outcomes}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", message = FALSE, warning = FALSE)
```

[`quantify()`](../reference/quantify.html) evaluates whether the **top event** occurs under each scenario. With the default `prob = FALSE`, inputs are **binary** (failed / not failed) for each basic event.

This vignette covers:

- Calling the boolean function from [`formulate()`](../reference/formulate.html) directly.
- Passing many scenarios as a data frame.
- Generating random binary scenarios from event probabilities (IT security tree).

## Setup: formulate the top event

```{r}
library(tidyfault)
library(dplyr)
library(purrr)
library(tibble)

data("db_nodes")
data("db_edges")

f <- curate(db_nodes, db_edges) |>
  equate() |>
  formulate()
sort(names(formals(f)))
```

---

## Single scenario: direct function call

Any scenario can be passed as named arguments matching `formals(f)`:

```{r}
f(DC = 1, AF = 1, SF = 0, BF = 0, NF = 0, AUF = 0, HF = 0, MF = 0)
```

---

## Many scenarios: `quantify()` on a data frame

Bundled [`db_outcomes_binary`](../reference/db_outcomes_binary.html) holds several rows of 0/1 inputs:

```{r}
data("db_outcomes_binary")
head(db_outcomes_binary)
```

[`quantify()`](../reference/quantify.html) returns one logical top-event outcome per row:

```{r}
mybinary <- quantify(f, newdata = db_outcomes_binary)
mybinary
```

---

## Batch evaluation: IT security tree

For larger batches, build a tibble of draws (one column per basic event), then call `quantify()` once. The example below uses [`it_security_probs`](../reference/it_security_probs.html) as Bernoulli parameters and fixes the seed for reproducibility.

```{r}
data("it_security_nodes")
data("it_security_edges")
data("it_security_probs")

f_sec <- formulate(equate(curate(it_security_nodes, it_security_edges)))

set.seed(123)
n_binary <- 5000
batch_binary <- it_security_probs |>
  map(~ rbinom(n_binary, size = 1, prob = .x)) |>
  as_tibble()

outcomes_batch <- quantify(f_sec, newdata = batch_binary)
table(outcomes_batch)
mean(outcomes_batch)
```

### Why vectorize?

- One call evaluates every row using the fast path when `fast = TRUE` (default).
- Column names must match `names(formals(f_sec))`.

---

## Summary

| Input | `quantify()` usage |
|-------|--------------------|
| One scenario | Named vector, list, or one-row data frame |
| Many scenarios | Data frame / tibble with one column per basic event |
| Top event type | Logical vector (`TRUE` = top event occurred) |

For probabilities instead of 0/1 inputs, see [quantify()-ing probabilities](quantify-probabilities.html).
