---
title: "quantify()-ing probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quantify()-ing probabilities}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

With `prob = TRUE`, [`quantify()`](../reference/quantify.html) treats each column as a **failure probability** for that basic event and returns the **top-event probability** for each row (assuming independence in the fast implementation).

This vignette shows:

1. Turning binary scenario tables into probability tables with [`populate()`](../reference/populate.html).
2. Projecting probabilities over time from hourly rates (IT security example).

## Database example: `populate()` then `quantify()`

[`db_probs`](../reference/db_probs.html) is in long form (`event`, `probability`). [`populate()`](../reference/populate.html) replaces 1s in a binary scenario with those probabilities (0 stays 0):

```{r}
library(tidyfault)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)

data("db_nodes")
data("db_edges")
data("db_outcomes_binary")
data("db_probs")

f <- curate(db_nodes, db_edges) |>
  equate() |>
  formulate()

db_outcomes_probs <- populate(db_outcomes_binary, db_probs)
knitr::kable(head(db_outcomes_probs, 6), digits = 3)
```

Then pass the probability matrix to `quantify()`:

```{r}
prob_top <- quantify(f, newdata = db_outcomes_probs, prob = TRUE)
prob_top
summary(prob_top)
```

---

## IT security: rates to cumulative probabilities over time

The next chunk defines illustrative constant failure rates per hour (same orders of magnitude as the JSS article). For each mission time `t`, map rates to **cumulative** failure probabilities with the exponential distribution $P(T \le t) = 1 - e^{-\lambda t}$:

```{r}
data("it_security_nodes")
data("it_security_edges")

f_sec <- formulate(equate(curate(it_security_nodes, it_security_edges)))

# Illustrative constant failure rates (per hour), as in the paper source
it_security_rates <- tibble::tibble(
  DA = 0.25 / 24 / 365,
  EP = 0.20 / 24 / 365,
  IM = 0.09 / 24 / 365,
  LR = 0.13 / 24 / 365,
  MN = 0.29 / 24 / 365,
  PO = 0.16 / 24 / 365,
  PC = 0.22 / 24 / 365,
  PM = 0.15 / 24 / 365,
  VS = 0.11 / 24 / 365,
  WB = 0.17 / 24 / 365
)
rate_names <- names(it_security_rates)

t_hours <- seq(0, 50000, by = 1000)
it_security_projections <- crossing(t = t_hours, it_security_rates) |>
  mutate(across(all_of(rate_names), ~ pexp(q = t, rate = .x)))

it_security_projections$outcome <- quantify(
  f_sec,
  newdata = it_security_projections[, rate_names, drop = FALSE],
  prob = TRUE
)

it_security_projections |>
  select(t, outcome) |>
  head(6)
```

Plot top-event probability versus time:

```{r}
ggplot(it_security_projections, aes(x = t, y = outcome)) +
  geom_line() +
  labs(
    x = "Time (hours)",
    y = "Top-event probability"
  )
```

---

## Summary

| Task | Approach |
|------|----------|
| Scenario-specific event probabilities | `populate()` from binary + long `event` / `probability` table |
| Direct probability scenarios | Data frame of probabilities + `quantify(..., prob = TRUE)` |
| Time profiles | Build rows of $P(\text{event})$ at each $t$, then same call |

Uncertainty simulation is covered in [Simulating uncertainty](simulating-uncertainty.html).
