---
title: "Visualizing fault trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing fault trees}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

**tidyfault** builds fault-tree plots on top of **ggplot2** and **ggraph**. The workhorse is [`illustrate()`](../reference/illustrate.html), which lays out nodes and edges; the [`plot()`](../reference/plot.html) method returns a ggplot object you can extend.

## Database example

Use `type = "all"` to draw gates and basic events together:

```{r}
library(tidyfault)
library(ggplot2)

data("db_nodes")
data("db_edges")

gg_db <- illustrate(nodes = db_nodes, edges = db_edges, type = "all")
plot(gg_db)
```

### What `illustrate()` returns

`illustrate()` returns a list with components used by `plot()` (node positions, edge paths, gate polygons). You can inspect `gg_db$nodes` or add ggplot layers on top of `plot(gg_db)`.

---

## IT security example (larger tree)

The bundled IT security tree ([`it_security_nodes`](../reference/it_security_nodes.html), [`it_security_edges`](../reference/it_security_edges.html)) uses `type = "both"` to emphasize gate structure and basic events:

```{r}
data("it_security_nodes")
data("it_security_edges")

gg_sec <- illustrate(
  nodes = it_security_nodes,
  edges = it_security_edges,
  type = "both"
)
plot(gg_sec)
```

---

## Workflow tip

Typical order:

1. Curate and analyze with the [core workflow](core-workflow.html) vignette.
2. Use **`illustrate()` → `plot()`** whenever you need to communicate structure to collaborators.

---

## Related reference

| Topic | Function |
|-------|----------|
| Layout and geometry | [`illustrate()`](../reference/illustrate.html) |
| ggplot wrapper | [`plot()`](../reference/plot.html) |
| Gate drawing primitives | [`gate_and()`](../reference/gate_and.html), [`gate_or()`](../reference/gate_or.html), [`gate_top()`](../reference/gate_top.html), [`get_gate()`](../reference/get_gate.html) |
